---
title: "Signal and Noise in [ILEARN [×]{.mult} WIDA ACCESS]{.nowrap}"
subtitle: "Where English-language proficiency starts to predict content-test performance in Indiana"
pagetitle: "Signal and Noise in ILEARN × WIDA ACCESS"
author:
  - name: "Damian W. Betebenner"
    affiliation: "National Center for the Improvement of Educational Assessment"
date: 2026-08-27
date-format: "[Indiana · ]MMMM D, YYYY"
format:
  revealjs:
    output-file: index.html
    theme: [default, styles/presentation-theme.scss]
    slide-number: true
    transition: fade
    transition-speed: fast
    width: 1920
    height: 1080
    margin: 0.06
    center: false
    hash: true
    history: true
    controls: true
    progress: true
    html-math-method: mathml
    fig-align: center
    embed-resources: true
    include-after-body:
      - text: |
          <script src="styles/presentation-figure-modal.js"></script>
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "svg"
    fig.align: "center"
---

```{r setup}
#| include: false

source("R/helpers.R")
R <- load_results()

n_pairs <- R$n$pairs_total
n_prim  <- R$n$pairs_primary
n_exit  <- R$n$exiters_primary
n_lag   <- R$n$lagged_primary

tau <- R$same_year$kendall_primary
lev <- as_dt(R$same_year$level_table_2025)
you <- as_dt(R$youden$by_stratum)
gcons <- as_dt(R$youden$by_grade_primary)
fx <- as_dt(R$first_crossing)
ex_ach <- as_dt(R$exiters$achievement_pooled)
ex_sgp <- as_dt(R$exiters$sgp_pooled)
ex_el  <- as_dt(R$exiters$el_status_agreement)
ex_n   <- as_dt(R$exiters$n_by_role_primary)
lag_p  <- as_dt(R$lagged_proficiency$pooled_primary)
lag_g  <- as_dt(R$lagged_proficiency$by_grade_summary)

## Copula-asymmetry summary (written by step 09; guarded for pre-09 renders).
asy <- tryCatch(data.table::fread("../outputs/asym_summary_by_year.csv"),
                error = function(e) NULL)
asy_tail <- function(y, d = "0.11") {
  if (!is.null(asy) && nrow(asy[YEAR == y])) fmt_num(asy[YEAR == y, tail_asymmetry][1], 2) else d
}
asy_exch <- if (!is.null(asy)) fmt_num(mean(asy$exch_index, na.rm = TRUE), 3) else "0.006"

lev_row <- function(label) {
  if (!nrow(lev)) return(list(n = NA, pctile = NA, p = NA, tau = NA))
  r <- lev[WIDA_LEVEL == label]
  if (!nrow(r)) return(list(n = NA, pctile = NA, p = NA, tau = NA))
  list(n = r$n[1], pctile = r$ilearn_pctile_median[1],
       p = r$p_ilearn_proficient[1], tau = r$kendall[1])
}
L1  <- lev_row("WIDA Level 1")
L2  <- lev_row("WIDA Level 2")
L43 <- lev_row("WIDA Level 4.3")
L5  <- lev_row("WIDA Level 5")

ex_band <- function(role, band) {
  if (!nrow(ex_ach)) return(NULL)
  if (is.na(band) || !nzchar(band)) {
    r <- ex_ach[role == role & (is.na(WIDA_EXIT_BAND) | WIDA_EXIT_BAND == "")]
  } else {
    r <- ex_ach[role == role & WIDA_EXIT_BAND == band]
  }
  if (!nrow(r)) return(NULL)
  r[1]
}

sgp_band <- function(role, band) {
  if (!nrow(ex_sgp)) return(NULL)
  r <- ex_sgp[role == role & WIDA_EXIT_BAND == band]
  if (!nrow(r)) return(NULL)
  r[1]
}
```

## Roadmap {.center}

::: {.visually-hidden}
Roadmap for the talk
:::

::: {style="font-size: 1.05em;"}
1. **The question** &mdash; is there a WIDA level where ILEARN stops looking like noise?
2. **Same-year signal** &mdash; location, not a jump in rank correlation; reading a cut with Youden's J
3. **The shape of the dependence** &mdash; the copula is asymmetric, and the asymmetry is the signal
4. **After ACCESS stops** &mdash; inferred exiters versus stayers
5. **A 50-50 proficiency cut** &mdash; run as a contrast, not a recommendation
6. **What this says about 4.3 and 5.0**
:::

::: {.key-insight}
**`r fmt_count(n_pairs)`** same-year ILEARN ELA × WIDA ACCESS Overall pairs after floor-score filters (**`r fmt_count(n_prim)`** in 2021–2025). WIDA is labeled **Overall Composite** throughout — the LONG stores it as `READING`.
:::

::: {.notes}
Sandbox analysis, not a client delivery and not the IDOE growth RFP. Every table is a testing-pattern or same-year pair — not an official exit roster.
:::


# The question {.center}

::: {.visually-hidden}
Section: the question
:::


## What we asked

States require ELA of every student, including English learners. WIDA ACCESS is the English-language proficiency test used to decide whether a student can access content in English.

The working claim:

- At low WIDA (roughly Levels 1–2), ILEARN ELA behaves like **noise**: floor pile-up, near-independence of the two ranks.
- At higher WIDA a **signal** appears: ILEARN distributions that start to look like they can access English-language content.

If that transition is real, it is evidence about where an exit criterion should sit. It is **not** itself the legal exit rule.

::: {.warning-box}
<p class="warning-box-title">Keep the construct straight</p>
WIDA ACCESS measures English language proficiency. ILEARN ELA measures grade-level content. Native English speakers fail ILEARN. A content-test cut is not a language-access cut.
:::


## Indiana's exit anchors

::: {.lead}
Two numbers define the current policy — everything that follows asks whether the data agree with them.
:::


Source: IDOE, *Exit Criteria for English Learners Guidance 2025–2026* (April 2025).

| Anchor | Meaning |
|---|---|
| **5.0** Overall composite | Absolute auto-exit |
| **4.3–4.9** Overall | Provisional pathway (grade 3+): ILP-committee additional evidence, not a pure scale-score cut |

WIDA LONG already stores L1–L4, **L4.3** (4.3–4.9 collapsed), L5, L6.

A fitted change-point describes where the content test starts to move with ACCESS. It may or may not equal the legal exit rule.


## Data and windows

Same-year, same-grade join: ILEARN `STUDENT_ID` = WIDA `ID`. Grades 3–8. Placeholder floors dropped (`ILEARN < 5000`, `WIDA == 100`).

::::: {.columns}
:::: {.column width="52%"}
| Window | Years | Role |
|---|---|---|
| **Primary** | 2021–2025 | Stable ILEARN scale. The estimand. |
| Ranks only | 2026 | Both scales reset. Percentiles, not proficiency. |
| Contrast | 2017–2019 | ISTEP / first ILEARN year. Not pooled. |

::: {.key-insight}
2026 ILEARN `SCALE_SCORE` in the combined LONG is the equated **old** scale. Ranks agree; the proficiency classifier does not.
:::
::::
:::: {.column width="48%"}
```{r}
#| output: asis
emit_html(fig_html("02_n_heatmap",
  "Paired N by year × grade. WIDA coverage in grades 3–8 is near-complete.",
  max_height = "520px"))
```
::::
:::::


# Same-year signal {.center}

::: {.visually-hidden}
Section: same-year dependence
:::


## The two tests are not independent

::: {.lead}
Before asking *where* a signal appears, the prior question is whether the two tests move together at all — which is what Kendall's τ measures (0 = independence, 1 = identical rank order).
:::

::::: {.columns}
:::: {.column width="44%"}
Primary-window Kendall's τ is **`r fmt_num(tau$kendall_min, 2)`–`r fmt_num(tau$kendall_max, 2)`** (median **`r fmt_num(tau$kendall_median, 2)`**) — strong in every grade and year.

The signal-versus-noise claim is therefore **local**, not global. What changes with WIDA level is **where** EL students sit on ILEARN, not whether the two ranks suddenly start moving together.
::::
:::: {.column width="56%"}
```{r}
#| output: asis
emit_html(fig_html("03_tau_heatmap",
  "Kendall's τ by grade and year; N pairs in parentheses. Click to enlarge.",
  max_height = "600px"))
```
::::
:::::


## The signal is location, not correlation

::: {.lead}
The core test of the hypothesis: as WIDA rises, does ILEARN rise by shifting the whole distribution upward, or merely by correlating more tightly *within* each band?
:::

| WIDA level | N | Within-level τ | Median ILEARN percentile | P(ILEARN proficient) |
|---|---:|---:|---:|---:|
| 1 | `r fmt_count(L1$n)` | `r fmt_tau(L1$tau)` | `r fmt_num(L1$pctile, 0)` | `r fmt_prob(L1$p)` |
| 2 | `r fmt_count(L2$n)` | `r fmt_tau(L2$tau)` | `r fmt_num(L2$pctile, 0)` | `r fmt_prob(L2$p)` |
| 4.3 (provisional) | `r fmt_count(L43$n)` | `r fmt_tau(L43$tau)` | `r fmt_num(L43$pctile, 0)` | `r fmt_prob(L43$p)` |
| 5 (auto-exit) | `r fmt_count(L5$n)` | `r fmt_tau(L5$tau)` | `r fmt_num(L5$pctile, 0)` | `r fmt_prob(L5$p)` |

::::: {.columns}
:::: {.column width="46%"}
**Location moves; correlation does not.** From L3 the ILEARN distribution walks up the grade — median percentile climbs from the high teens at L1 to the low 90s at L5 — while the **within-band τ stays modest** the whole way (restriction of range). The signal is *where* students sit, not a tightening rank correlation.
::::
:::: {.column width="54%"}
```{r}
#| output: asis
emit_html(fig_html("04_ilearn_by_level_2025",
  "ILEARN ELA location by WIDA ACCESS Overall level, 2025.",
  max_height = "540px"))
```
::::
:::::


## Reading a WIDA cut as a classifier

Any candidate cut turns WIDA ACCESS into a yes/no test: **WIDA ≥ cut** predicts "ILEARN-proficient."

::::: {.columns}
:::: {.column width="44%"}
- **Sensitivity** — of the ILEARN-proficient students, the share the cut catches (true-positive rate).
- **Specificity** — of the not-proficient students, the share correctly left below (true-negative rate).

**Youden's J = sensitivity + specificity − 1.** J runs from 0 (a coin flip) to 1 (perfect separation); on an ROC curve it is the vertical height above the chance diagonal. The **J-optimal cut** best separates the two groups.
::::
:::: {.column width="56%"}
```{r}
#| output: asis
emit_html(fig_html("05b_youden_explainer",
  "Youden's J: the cut that maximizes sensitivity + specificity − 1.",
  max_height = "560px"))
```
::::
:::::


## Youden and the exit cut

The J-optimal WIDA cut is the score that best sorts EL students into ILEARN-proficient vs not. It is a **descriptive** answer — where the content signal separates best — not a legal rule.

- Across 2021–2025 the J-optimal cut against ILEARN proficiency sits near **4.1** (grade medians 3.9–4.6): inside the **4.3** committee band, never at **5.0**.
- At **5.0** the proficient EL students below the cut are so few that **sensitivity collapses** — the cut misses most of the students it should flag, so J falls.
- Youden weights a false exit and a false hold **equally**. A policy that treats an early exit as the costlier error would rationally sit *above* the J-optimal cut — one reading of why 5.0 is defensible even though J prefers 4.3.

::: {.warning-box}
<p class="warning-box-title">Optimal ≠ correct construct</p>
J optimizes agreement with **ILEARN proficiency** — a content outcome, not an English-access outcome. A statistically optimal classifier of the wrong target is still the wrong exit rule.
:::


## The best statistical cut lands near 4.3, not 5.0

::: {.lead}
If we let the data choose the WIDA cut that best separates ILEARN-proficient students from the rest, where does it land relative to the 4.3 and 5.0 anchors?
:::


Primary-window Youden-optimal WIDA cuts (same-year proficiency) have median **`r if (nrow(you[YEAR_STRATUM=="primary"])) fmt_num(you[YEAR_STRATUM=="primary", best_cut_median], 2) else "—"`**.

```{r}
#| results: asis
if (nrow(gcons)) {
  knitr::kable(gcons[, .(Grade = GRADE,
      `Median optimal cut` = sprintf("%.1f", best_cut_median),
      SD = sprintf("%.2f", best_cut_sd))],
    format = "html", align = "lrr")
}
```

Grades 4–5 sit near the start of the **4.3** band; 3 and 6–8 sit a little below it. None of the Youden optima are at **5.0**. That is not an argument that 5.0 is the wrong legal exit — ILEARN proficiency is a high bar, and even students first reaching 5.0 are only about **`r if (nrow(fx[cut==5])) fmt_prob(fx[cut==5, p_proficient_at]) else "0.40"`** ILEARN-proficient that year.


## Most of the climb happens before auto-exit

::: {.lead}
Following students the first year they reach a cut *while still tested*: how high are they already sitting on ILEARN by 4.3 versus 5.0?
:::


Students first reaching a cut **while they still have a WIDA score**:

| Cut | N | ILEARN percentile before → at | P(proficient) before → at |
|---|---:|---|---|
| 4.3 | `r if (nrow(fx[cut==4.3])) fmt_count(fx[cut==4.3, n]) else "—"` | `r if (nrow(fx[cut==4.3])) paste(fmt_num(fx[cut==4.3, ilearn_pctile_prior], 0), "→", fmt_num(fx[cut==4.3, ilearn_pctile_at], 0)) else "—"` | `r if (nrow(fx[cut==4.3])) paste(fmt_prob(fx[cut==4.3, p_proficient_prior]), "→", fmt_prob(fx[cut==4.3, p_proficient_at])) else "—"` |
| 5.0 | `r if (nrow(fx[cut==5])) fmt_count(fx[cut==5, n]) else "—"` | `r if (nrow(fx[cut==5])) paste(fmt_num(fx[cut==5, ilearn_pctile_prior], 0), "→", fmt_num(fx[cut==5, ilearn_pctile_at], 0)) else "—"` | `r if (nrow(fx[cut==5])) paste(fmt_prob(fx[cut==5, p_proficient_prior]), "→", fmt_prob(fx[cut==5, p_proficient_at])) else "—"` |

Most of the ILEARN location gain has already happened by the time a student hits auto-exit. This is still the **stayer** sample.


# The shape of the dependence {.center}

::: {.visually-hidden}
Section: copula asymmetry
:::


## Why symmetric copulas hide the story

The confirmatory families in step 03 — **t, Frank, Gaussian** — are all *exchangeable* and *radially symmetric*. Two consequences, both by construction:

- They force lower-tail dependence to equal upper-tail dependence: $\lambda_L = \lambda_U$.
- They treat the lower-left corner (both tests **low** — the WIDA floor) as a mirror of the upper-right corner (both **high**).

But the signal-vs-noise hypothesis is exactly the claim that those corners are **not** mirror images: low WIDA is noise, high WIDA is signal. A radially symmetric copula cannot represent that — so the departure from symmetry is not a nuisance to smooth away. It **is** the hypothesis, written in dependence-structure terms.

::: {.key-insight}
Measured directly, the copula is **exchangeable** (swapping the two tests barely moves it; index ≈ `r asy_exch`) but **radially asymmetric**.
:::


## A picture of the asymmetry

::: {.lead}
If the dependence were radially symmetric this map would be flat grey — here is what it actually is.
:::

::::: {.columns}
:::: {.column width="42%"}
The both-high **signal corner** and the both-low **noise corner** are **not mirror images** — the surface is far from flat, and the coupling is pulled toward high WIDA.

Blue marks where high-scorer pairs cluster more tightly than low-scorer pairs — structure the symmetric t / Frank / Gaussian families cannot represent.
::::
:::: {.column width="58%"}
```{r}
#| output: asis
emit_html(fig_html("09_copula_asymmetry_surface",
  "D(u,v) = empirical copula minus its reflection, primary window.",
  max_height = "640px"))
```
::::
:::::


## The signal corner couples more than the noise corner

::: {.lead}
Another view of the same fact: how tightly the two tests co-move among the lowest scorers versus the highest.
:::

::::: {.columns}
:::: {.column width="44%"}
Deep in the tail, upper-tail concordance $\lambda_U$ runs well above lower-tail $\lambda_L$: **both-high pairs co-occur more than both-low pairs.** The gap $\lambda_U-\lambda_L$ grows from **`r asy_tail(2021)` (2021)** to **`r asy_tail(2025, "0.22")` (2025)** — stable in sign, strengthening in size, and invisible to any t / Frank / Gaussian fit.

A second, independent signature of the same access story.
::::
:::: {.column width="56%"}
```{r}
#| output: asis
emit_html(fig_html("09_copula_asymmetry_tails",
  "Upper- vs lower-tail dependence functions, primary window. The shaded gap is the asymmetry.",
  max_height = "520px"))
```
::::
:::::


# After ACCESS stops {.center}

::: {.visually-hidden}
Section: inferred exiters
:::


## Testing-pattern exit

::: {.lead}
With no official exit roster in these files, we infer exit from the one pattern both datasets share — who stops taking ACCESS.
:::


Exit is the complementary event: ILEARN + WIDA at *t*, ILEARN only at *t+1*.

| Role | Pattern |
|---|---|
| **Exiter** | Both tests at *(t, g)*; ILEARN at *(t+1, g+1)*; no WIDA at *t+1* |
| **Stayer** | Both tests both years |
| **Never-EL** | ILEARN both years, no WIDA, not classified EL |
| **Attritor** | Both tests at *t*, no ILEARN at *t+1* (includes all grade 8) |

::: {.warning-box}
<p class="warning-box-title">Not an official roster</p>
These LONGs do not carry IDOE’s exit list. `<4.3` disappearances are documented, not treated as policy exits. Grade 8 cannot be followed.
:::

::: {.key-insight}
**`r fmt_count(n_exit)`** inferred exiters in the primary window (grades 3–7).
:::


## Who actually leaves ACCESS

::: {.lead}
Turning to the event an exit rule actually governs: from which WIDA band do students leave ACCESS, and how has that shifted over five years?
:::

::::: {.columns}
:::: {.column width="40%"}
- Almost every inferred exit comes from **≥5.0** — auto-exit is doing the work.
- The **4.3–4.9** committee band barely fills in until **2024**, then jumps.
- **<4.3** disappearances stay in the low hundreds every year — mobility or a missed test, not a policy pathway.
::::
:::: {.column width="60%"}
```{r}
#| output: asis
emit_html(fig_html("07_exiter_n_heatmap",
  "Inferred exiters by year × last WIDA ACCESS Overall band.",
  max_height = "560px"))
```
::::
:::::


## The classification flag agrees only at 5.0

::: {.lead}
A testing-pattern “exit” only means something if it matches an official reclassification — does the EL-status flag agree these students left?
:::


Share of inferred exiters whose ILEARN `EL_STATUS` moves LEP → Not an English Language Learner:

```{r}
#| results: asis
if (nrow(ex_el)) {
  knitr::kable(ex_el[, .(
    `Last WIDA` = WIDA_EXIT_BAND,
    N = prettyNum(n, big.mark = ","),
    `P(LEP → not EL)` = paste(fmt_prob(p_el_to_not_el_min), "–", fmt_prob(p_el_to_not_el_max)),
    `P(still EL)` = paste(fmt_prob(p_still_el_min), "–", fmt_prob(p_still_el_max))
  )], format = "html", align = "lrrr")
}
```

`>=5.0` testing-pattern matches the official classification. `4.3–4.9` and `<4.3` mostly do **not** — ACCESS non-takers, still classified EL.


## No post-exit collapse

::: {.lead}
The central fear about exiting a student is academic collapse once services stop — do ACCESS leavers fall on ILEARN the next year?
:::

::::: {.columns}
:::: {.column width="42%"}
No. **≥5.0** leavers sit ~9–10 percentile points **above** the never-EL median and hold or rise. **4.3–4.9** leavers sit near never-EL and dip about 2 points. **<4.3** disappearances stay on the ILEARN floor — like the students who never left.
::::
:::: {.column width="58%"}
```{r}
#| output: asis
emit_html(fig_html("07_exiter_percentile",
  "ILEARN ELA percentile at t (still tested) vs t+1 (ILEARN only).",
  max_height = "440px"))
```
::::
:::::

```{r}
#| results: asis
if (nrow(ex_ach[role == "exiter"])) {
  knitr::kable(ex_ach[role == "exiter", .(
    Band = WIDA_EXIT_BAND,
    N = prettyNum(n, big.mark = ","),
    `Pctile t` = sprintf("%.0f", ilearn_pctile_t),
    `Pctile t+1` = sprintf("%.0f", ilearn_pctile_t1),
    `Δ` = sprintf("%+.0f", ilearn_pctile_delta),
    `P(prof) t` = fmt_prob(p_proficient_t),
    `P(prof) t+1` = fmt_prob(p_proficient_t1)
  )], format = "html", align = "lrrrrrr")
}
```


## Exit-year growth sits at or above 50

::: {.lead}
Beyond where exiters *stand*, how do they *grow* in their first year without ACCESS, relative to typical students?
:::

::::: {.columns}
:::: {.column width="42%"}
ILEARN SGP at *t+1* is growth in the first year without ACCESS. **≥5.0** leavers post a median **58** — a right shift, not a left one. **4.3–4.9** leavers post ~**50**. Only **<4.3** disappearances are left-shifted (median **41**), which is what an unexpected drop among still-low students looks like.
::::
:::: {.column width="58%"}
```{r}
#| output: asis
emit_html(fig_html("07_exiter_sgp",
  "Median ILEARN SGP in the exit year. Faceted by last WIDA band.",
  max_height = "440px"))
```
::::
:::::

```{r}
#| results: asis
if (nrow(ex_sgp[role == "exiter"])) {
  knitr::kable(ex_sgp[role == "exiter", .(
    Band = WIDA_EXIT_BAND,
    N = prettyNum(n, big.mark = ","),
    `Median SGP` = sprintf("%.0f", sgp_median),
    `P(SGP < 35)` = fmt_prob(p_sgp_below_35),
    `P(SGP > 65)` = fmt_prob(p_sgp_above_65)
  )], format = "html", align = "lrrrr")
}
```


# A 50-50 proficiency cut {.center}

::: {.visually-hidden}
Section: lagged copula contrast
:::


## The recipe that gets proposed

Find the WIDA ACCESS Overall score such that a student has better than a **50-50 chance** of being ILEARN ELA-proficient **the following year**.

Read the joint through a copula $C(U_t^{\mathrm{WIDA}}, V_{t+1}^{\mathrm{ILEARN}})$. Let $v^*$ be the ILEARN proficiency cut on the paired rank scale. Invert

$$
P(V \ge v^* \mid U = u) = \tfrac{1}{2}
$$

and map $u^*$ back to a WIDA proficiency level.

Also report the empirical local-window version: the first cut $c$ at which $P(\text{proficient}_{t+1} \mid \mathrm{WIDA}_t \approx c) \ge 0.5$.

::: {.key-insight}
**`r fmt_count(n_lag)`** lagged pairs (WIDA at *t*, ILEARN at *t+1*) in the primary window.
:::


## Why we do not recommend it

::: {.warning-box}
<p class="warning-box-title">Wrong construct</p>
ILEARN proficiency is not a proxy for being ready to leave EL services. Never-EL students — native and fluent English speakers who never took ACCESS — are **not** all ILEARN-proficient. In the primary window their P(proficient) is about **`r if (nrow(lag_p)) fmt_prob(lag_p$p_never_el_proficient[1]) else "0.43"`**.
:::

If the bar is “50-50 on next year’s ELA test,” a substantial share of never-EL students fail it. Using that bar to decide English-learner status asks ACCESS to predict a content outcome that English itself does not guarantee.

We still run the recipe. It is part of the conversation. It is filed as a **contrast**.


## What the copula actually returns

::: {.lead}
Running the 50-50 recipe honestly: the modeled probability of next-year ILEARN proficiency as a function of this year's WIDA score.
:::

::::: {.columns}
:::: {.column width="42%"}
The copula conditional (solid) and the local empirical rate (points) both rise with WIDA and cross the **0.50** line only near the top of the scale. The dotted line is the never-EL rate — the bar even fluent English speakers do not all clear.
::::
:::: {.column width="58%"}
```{r}
#| output: asis
emit_html(fig_html("08_lagged_p_prof_primary",
  "P(ILEARN proficient at t+1 | WIDA Overall at t). Solid = copula; points = local empirical; 0.50 line; dotted = never-EL.",
  max_height = "540px"))
```
::::
:::::


## Where the 50-50 cut sits

::: {.lead}
Reading the cut off that curve and setting it against the policy anchors — and against the fact that not even all never-EL students clear this bar.
:::

::::: {.columns}
:::: {.column width="52%"}
```{r}
#| results: asis
if (nrow(lag_p)) {
  knitr::kable(data.frame(
    Quantity = c(
      "Pooled copula cut (WIDA PL at t)",
      "Pooled local-empirical cut",
      "P(prof t+1 | WIDA ≥ 4.3)",
      "P(prof t+1 | WIDA ≥ 5.0)",
      "Local P(prof t+1 | WIDA ≈ 4.3)",
      "Local P(prof t+1 | WIDA ≈ 5.0)",
      "Never-EL P(ILEARN proficient)"
    ),
    Value = c(
      fmt_num(lag_p$copula_cut[1], 2),
      fmt_num(lag_p$empirical_local_cut[1], 2),
      fmt_prob(lag_p$p_ge_4.3[1]),
      fmt_prob(lag_p$p_ge_5.0[1]),
      fmt_prob(lag_p$p_local_4.3[1]),
      fmt_prob(lag_p$p_local_5.0[1]),
      fmt_prob(lag_p$p_never_el_proficient[1])
    )
  ), format = "html", align = "lr", row.names = FALSE)
}
```
::::
:::: {.column width="48%"}
```{r}
#| output: asis
emit_html(fig_html("08_lagged_fifty_fifty_by_grade",
  "Fifty-fifty WIDA cut by ILEARN grade. Copula vs local empirical. Dashed = 4.3 and 5.0.",
  max_height = "460px"))
```
::::
:::::

::: {.notes}
If the cut lands near 5.0, the recipe is rediscovering auto-exit by asking ACCESS to predict a hard content outcome. If it lands near 4.3, it agrees with Youden / the location table — still not a reason to treat ILEARN proficiency as the exit construct.
:::


# 4.3 versus 5.0 {.center}

::: {.visually-hidden}
Section: what this says about the exit threshold
:::


## How the pieces fit

1. ILEARN ELA is not informative about English access below WIDA **Level 3**. L1–L2 is noise in the sense of the hypothesis.
2. Signal is clearly present by the **4.3–4.9** provisional band. That is also where Youden against ILEARN proficiency peaks.
3. **5.0 auto-exit** selects students already at the top of the ILEARN distribution. It is a conservative access standard, not the first point at which ILEARN stops looking like noise.
4. Students who leave ACCESS from **5.0** hold their ILEARN location and post slightly-above-typical growth. The classification flag agrees they exited.
5. Students who disappear from ACCESS at **4.3–4.9** also hold location — but most are still classified EL. The committee-band “exiter” cell is thin and contaminated.
6. The 50-50 next-year-proficiency recipe is in the JSON and the figures. It is **not** the recommended exit rule.

::: {.key-insight}
On these files the threshold debate is about **who is allowed to leave**, not about academic collapse after ACCESS stops.
:::


## Reproducible artifacts

::: {.lead}
Every number in this deck is regenerable from two read-only source files, and no student-level record ever leaves the pipeline.
:::

::::: {.columns}
:::: {.column width="52%"}
**One source, many channels.** The R pipeline (steps 00–09) is the single source of truth; the tables, JSON manifests, figures, and this deck are all *derived* from it and rebuilt by one command — so a corrected function or a new year of data propagates everywhere without hand-editing.

**Containment / FERPA.** Source LONGs are read-only and never copied in. Student-level pairs stay in gitignored `cache/`. Committed outputs are aggregates only: `write_output()` refuses `ID` / `STUDENT_ID` columns, and any cell below N = 10 is dropped.
::::
:::: {.column width="48%"}
| Layer | Where |
|---|---|
| Analysis | `00–09_*.R`, `R/`, `config.R` |
| Tables | `outputs/*.csv` |
| JSON manifests | `outputs/manifests/*.json` |
| Figures | `figures/*.png` (+ PDF) |
| This deck | `docs/index.html` |

```bash
Rscript run_all.R    # rebuild 00–09 + manifests
./docs/render.sh     # re-render this deck
```
::::
:::::

::: {.notes}
Pages source: Deploy from a branch → /docs. embed-resources is on, so the HTML is self-contained.
:::


# Where this could go next {.center}

::: {.visually-hidden}
Section: future investigations
:::


## Open the black box: WIDA's four domains

::: {.lead}
Everything here used the **Overall** composite — but WIDA scores four modalities, and English *content* access is largely a literacy story.
:::

The same rank / copula / threshold machinery, run on **Listening, Speaking, Reading, and Writing** separately (pulling the domain subscores from the ACCESS source files), would ask *which* skill the ILEARN signal is actually attached to.

- Hypothesis: **Reading** and **Writing** carry most of the ILEARN-ELA dependence; **Listening** and **Speaking** far less.
- A **vine copula** on the four domains plus ILEARN would isolate the content signal *conditionally* — the dependence that survives after the other domains are held fixed.

::: {.key-insight}
**Would reveal:** whether a defensible exit rule should weight the literacy domains above the oral ones, rather than collapsing all four into a single Overall number.
:::


## Is the low-WIDA "noise" real — or measurement error?

::: {.lead}
At Levels 1–2 the two ranks barely move together; the open question is whether that is genuine independence or just unreliability where both tests sit near their floors.
:::

- **Disattenuate** the τ and the copula using each test's conditional standard errors of measurement (CSEMs), where the source files carry them — noise from unreliability is not the same as noise from access.
- Model the ILEARN floor explicitly with a **censored / two-part (Tobit-style)** specification, separating "cannot yet access the test" from "low measured ability."

::: {.key-insight}
**Would reveal:** how much of the L1–L2 "noise" is a substantive claim about English access versus an artifact of thin measurement information at the floor — which sets how much weight that region can bear.
:::


## Direction and dynamics

::: {.lead}
The cross-sectional picture cannot say whether language *gain* precedes content *gain* — but the multi-year panel can.
:::

- A **random-intercept cross-lagged panel model** (RI-CLPM) on WIDA$_t \rightarrow$ ILEARN$_{t+1}$ versus ILEARN$_t \rightarrow$ WIDA$_{t+1}$ tests the directional ordering of the access relationship.
- Add **time in program** (years since a student's first ACCESS administration, recoverable from the panel) as a moderator: does the WIDA→ILEARN coupling strengthen with exposure?

::: {.key-insight}
**Would reveal:** whether rising English proficiency plausibly *drives* content access — the theory of action an exit rule assumes — or merely co-occurs with it.
:::


## Beyond the median, beyond one cut

::: {.lead}
We summarized each WIDA level by its median and chose a single optimal cut; both hide heterogeneity that matters for who a rule serves.
:::

- **Quantile / distributional regression** of ILEARN on WIDA PL models the *whole* conditional distribution — where the lower tail crosses proficiency, not just the center.
- **Finite-mixture / latent-class copulas** ask whether newcomers, long-term ELs, and students with interrupted schooling follow *different* WIDA→ILEARN maps that a single cut averages away.
- A **Bayesian or bootstrapped interval** on the τ(u) change-point would give the transition "region" an honest width.

::: {.key-insight}
**Would reveal:** which students a single Overall cut serves well, and which it systematically mis-sorts.
:::


## From a threshold to a decision rule

::: {.lead}
A statistically optimal cut is not yet a policy; two extensions turn description into a defensible decision.
:::

- **Cost-weighted Youden:** recompute the optimal cut as a function of the false-exit / false-hold cost ratio, making the value judgment explicit instead of fixing it implicitly at equal weights.
- **Multistate / survival modeling** of exit → reclassification → long-term-EL risk: what happens *years* after a student leaves, not only at $t{+}1$.
- **Subgroup invariance:** does the WIDA→ILEARN relationship hold across home language, grade of entry, and disability status, or does one cut carry hidden differential impact?

::: {.key-insight}
**Would reveal:** a cut chosen with stated priorities, audited for long-run consequences and equity — the form an actual recommendation would have to take.
:::


## The through-line

::: {.lead}
Each next step sharpens the same question this deck opened with: where does a language test stop measuring language and start gating content?
:::

The domain decomposition asks *which* language skill gates access; the measurement-error and distributional work ask *how sharply* and *for whom*; the panel and decision-analytic work ask *in what direction* and *at what cost*. None of them replaces the ILP committee — they give it a better-instrumented map of the band where the content test becomes a content test.

