---
title: "ILEARN × WIDA ACCESS"
subtitle: "Where the content test becomes a content test"
author:
  - name: "Damian W. Betebenner"
    affiliation: "National Center for the Improvement of Educational Assessment"
date: 2026-08-27
date-format: "[Indiana · ]MMMM D, YYYY"
format:
  revealjs:
    output-file: index.html
    theme: [default, styles/presentation-theme.scss]
    slide-number: true
    transition: fade
    transition-speed: fast
    width: 1920
    height: 1080
    margin: 0.06
    center: false
    hash: true
    history: true
    controls: true
    progress: true
    html-math-method: mathml
    fig-align: center
    embed-resources: true
    include-after-body:
      - text: |
          <script src="styles/presentation-figure-modal.js"></script>
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "svg"
    fig.align: "center"
---

```{r setup}
#| include: false

source("R/helpers.R")
R <- load_results()

n_pairs <- R$n$pairs_total
n_prim  <- R$n$pairs_primary
n_exit  <- R$n$exiters_primary
n_lag   <- R$n$lagged_primary

tau <- R$same_year$kendall_primary
lev <- as_dt(R$same_year$level_table_2025)
you <- as_dt(R$youden$by_stratum)
gcons <- as_dt(R$youden$by_grade_primary)
fx <- as_dt(R$first_crossing)
ex_ach <- as_dt(R$exiters$achievement_pooled)
ex_sgp <- as_dt(R$exiters$sgp_pooled)
ex_el  <- as_dt(R$exiters$el_status_agreement)
ex_n   <- as_dt(R$exiters$n_by_role_primary)
lag_p  <- as_dt(R$lagged_proficiency$pooled_primary)
lag_g  <- as_dt(R$lagged_proficiency$by_grade_summary)

lev_row <- function(label) {
  if (!nrow(lev)) return(list(n = NA, pctile = NA, p = NA, tau = NA))
  r <- lev[WIDA_LEVEL == label]
  if (!nrow(r)) return(list(n = NA, pctile = NA, p = NA, tau = NA))
  list(n = r$n[1], pctile = r$ilearn_pctile_median[1],
       p = r$p_ilearn_proficient[1], tau = r$kendall[1])
}
L1  <- lev_row("WIDA Level 1")
L2  <- lev_row("WIDA Level 2")
L43 <- lev_row("WIDA Level 4.3")
L5  <- lev_row("WIDA Level 5")

ex_band <- function(role, band) {
  if (!nrow(ex_ach)) return(NULL)
  if (is.na(band) || !nzchar(band)) {
    r <- ex_ach[role == role & (is.na(WIDA_EXIT_BAND) | WIDA_EXIT_BAND == "")]
  } else {
    r <- ex_ach[role == role & WIDA_EXIT_BAND == band]
  }
  if (!nrow(r)) return(NULL)
  r[1]
}

sgp_band <- function(role, band) {
  if (!nrow(ex_sgp)) return(NULL)
  r <- ex_sgp[role == role & WIDA_EXIT_BAND == band]
  if (!nrow(r)) return(NULL)
  r[1]
}
```

## Roadmap {.center}

::: {.visually-hidden}
Roadmap for the talk
:::

::: {style="font-size: 1.05em;"}
1. **The question** &mdash; is there a WIDA level where ILEARN stops looking like noise?
2. **Same-year signal** &mdash; location, not a jump in rank correlation
3. **After ACCESS stops** &mdash; inferred exiters versus stayers
4. **A 50-50 proficiency cut** &mdash; run as a contrast, not a recommendation
5. **What this says about 4.3 and 5.0**
:::

::: {.key-insight}
**`r fmt_count(n_pairs)`** same-year ILEARN ELA × WIDA ACCESS Overall pairs after floor-score filters (**`r fmt_count(n_prim)`** in 2021–2025). WIDA is labeled **Overall Composite** throughout — the LONG stores it as `READING`.
:::

::: {.notes}
Sandbox analysis, not a client delivery and not the IDOE growth RFP. Every table is a testing-pattern or same-year pair — not an official exit roster.
:::


# The question {.center}

::: {.visually-hidden}
Section: the question
:::


## What we asked

States require ELA of every student, including English learners. WIDA ACCESS is the English-language proficiency test used to decide whether a student can access content in English.

The working claim:

- At low WIDA (roughly Levels 1–2), ILEARN ELA behaves like **noise**: floor pile-up, near-independence of the two ranks.
- At higher WIDA a **signal** appears: ILEARN distributions that start to look like they can access English-language content.

If that transition is real, it is evidence about where an exit criterion should sit. It is **not** itself the legal exit rule.

::: {.warning-box}
<p class="warning-box-title">Keep the construct straight</p>
WIDA ACCESS measures English language proficiency. ILEARN ELA measures grade-level content. Native English speakers fail ILEARN. A content-test cut is not a language-access cut.
:::


## Indiana's exit anchors

Source: IDOE, *Exit Criteria for English Learners Guidance 2025–2026* (April 2025).

| Anchor | Meaning |
|---|---|
| **5.0** Overall composite | Absolute auto-exit |
| **4.3–4.9** Overall | Provisional pathway (grade 3+): ILP-committee additional evidence, not a pure scale-score cut |

WIDA LONG already stores L1–L4, **L4.3** (4.3–4.9 collapsed), L5, L6.

A fitted change-point describes where the content test starts to move with ACCESS. It may or may not equal the legal exit rule.


## Data and windows

Same-year, same-grade join: ILEARN `STUDENT_ID` = WIDA `ID`. Grades 3–8. Placeholder floors dropped (`ILEARN < 5000`, `WIDA == 100`).

| Window | Years | Role |
|---|---|---|
| **Primary** | 2021–2025 | Stable ILEARN scale. The estimand. |
| Ranks only | 2026 | Both scales reset. Percentiles, not proficiency. |
| Contrast | 2017–2019 | ISTEP / first ILEARN year. Not pooled. |

::: {.key-insight}
2026 ILEARN `SCALE_SCORE` in the combined LONG is the equated **old** scale. Native 2026 medians are a different test. Ranks agree; the proficiency classifier does not.
:::


# Same-year signal {.center}

::: {.visually-hidden}
Section: same-year dependence
:::


## The two tests are not independent

Primary-window Kendall’s τ is **`r fmt_num(tau$kendall_min, 2)`–`r fmt_num(tau$kendall_max, 2)`** (median **`r fmt_num(tau$kendall_median, 2)`**).

The signal-versus-noise claim is therefore **local**, not global. Overall rank dependence is already strong. What changes with WIDA level is **where** EL students sit on ILEARN — not whether the two ranks suddenly start moving together.

```{r}
#| output: asis
emit_html(fig_html("02_n_heatmap",
  "Paired N by year × grade. Click to enlarge.",
  max_height = "520px"))
```


## 2025: location, not a τ jump

| WIDA level | N | Within-level τ | Median ILEARN percentile | P(ILEARN proficient) |
|---|---:|---:|---:|---:|
| 1 | `r fmt_count(L1$n)` | `r fmt_tau(L1$tau)` | `r fmt_num(L1$pctile, 0)` | `r fmt_prob(L1$p)` |
| 2 | `r fmt_count(L2$n)` | `r fmt_tau(L2$tau)` | `r fmt_num(L2$pctile, 0)` | `r fmt_prob(L2$p)` |
| 4.3 (provisional) | `r fmt_count(L43$n)` | `r fmt_tau(L43$tau)` | `r fmt_num(L43$pctile, 0)` | `r fmt_prob(L43$p)` |
| 5 (auto-exit) | `r fmt_count(L5$n)` | `r fmt_tau(L5$tau)` | `r fmt_num(L5$pctile, 0)` | `r fmt_prob(L5$p)` |

L1–L2: ILEARN is on the floor. From L3 the distribution walks up the grade. By L4.3 paired EL students sit high in the grade; by L5 they sit at the top. Within-band τ stays modest the whole way (restriction of range).

```{r}
#| output: asis
emit_html(fig_html("04_ilearn_by_level_2025",
  "ILEARN ELA location by WIDA ACCESS Overall level, 2025.",
  max_height = "380px"))
```


## Youden against ILEARN proficiency is not 5.0

Primary-window Youden-optimal WIDA cuts (same-year proficiency) have median **`r if (nrow(you[YEAR_STRATUM=="primary"])) fmt_num(you[YEAR_STRATUM=="primary", best_cut_median], 2) else "—"`**.

```{r}
#| results: asis
if (nrow(gcons)) {
  knitr::kable(gcons[, .(Grade = GRADE,
      `Median optimal cut` = sprintf("%.1f", best_cut_median),
      SD = sprintf("%.2f", best_cut_sd))],
    format = "html", align = "lrr")
}
```

Grades 4–5 sit near the start of the **4.3** band; 3 and 6–8 sit a little below it. None of the Youden optima are at **5.0**. That is not an argument that 5.0 is the wrong legal exit — ILEARN proficiency is a high bar, and even students first reaching 5.0 are only about **`r if (nrow(fx[cut==5])) fmt_prob(fx[cut==5, p_proficient_at]) else "0.40"`** ILEARN-proficient that year.


## First crossing — still taking ACCESS

Students first reaching a cut **while they still have a WIDA score**:

| Cut | N | ILEARN percentile before → at | P(proficient) before → at |
|---|---:|---|---|
| 4.3 | `r if (nrow(fx[cut==4.3])) fmt_count(fx[cut==4.3, n]) else "—"` | `r if (nrow(fx[cut==4.3])) paste(fmt_num(fx[cut==4.3, ilearn_pctile_prior], 0), "→", fmt_num(fx[cut==4.3, ilearn_pctile_at], 0)) else "—"` | `r if (nrow(fx[cut==4.3])) paste(fmt_prob(fx[cut==4.3, p_proficient_prior]), "→", fmt_prob(fx[cut==4.3, p_proficient_at])) else "—"` |
| 5.0 | `r if (nrow(fx[cut==5])) fmt_count(fx[cut==5, n]) else "—"` | `r if (nrow(fx[cut==5])) paste(fmt_num(fx[cut==5, ilearn_pctile_prior], 0), "→", fmt_num(fx[cut==5, ilearn_pctile_at], 0)) else "—"` | `r if (nrow(fx[cut==5])) paste(fmt_prob(fx[cut==5, p_proficient_prior]), "→", fmt_prob(fx[cut==5, p_proficient_at])) else "—"` |

Most of the ILEARN location gain has already happened by the time a student hits auto-exit. This is still the **stayer** sample.


# After ACCESS stops {.center}

::: {.visually-hidden}
Section: inferred exiters
:::


## Testing-pattern exit

Exit is the complementary event: ILEARN + WIDA at *t*, ILEARN only at *t+1*.

| Role | Pattern |
|---|---|
| **Exiter** | Both tests at *(t, g)*; ILEARN at *(t+1, g+1)*; no WIDA at *t+1* |
| **Stayer** | Both tests both years |
| **Never-EL** | ILEARN both years, no WIDA, not classified EL |
| **Attritor** | Both tests at *t*, no ILEARN at *t+1* (includes all grade 8) |

::: {.warning-box}
<p class="warning-box-title">Not an official roster</p>
These LONGs do not carry IDOE’s exit list. `<4.3` disappearances are documented, not treated as policy exits. Grade 8 cannot be followed.
:::

::: {.key-insight}
**`r fmt_count(n_exit)`** inferred exiters in the primary window (grades 3–7).
:::


## Who actually leaves ACCESS

```{r}
#| output: asis
emit_html(fig_html("07_exiter_n_heatmap",
  "Inferred exiters by year × last WIDA ACCESS Overall band. Almost all leave from ≥5.0 until 2024.",
  max_height = "560px"))
```


## The classification flag agrees only at 5.0

Share of inferred exiters whose ILEARN `EL_STATUS` moves LEP → Not an English Language Learner:

```{r}
#| results: asis
if (nrow(ex_el)) {
  knitr::kable(ex_el[, .(
    `Last WIDA` = WIDA_EXIT_BAND,
    N = prettyNum(n, big.mark = ","),
    `P(LEP → not EL)` = paste(fmt_prob(p_el_to_not_el_min), "–", fmt_prob(p_el_to_not_el_max)),
    `P(still EL)` = paste(fmt_prob(p_still_el_min), "–", fmt_prob(p_still_el_max))
  )], format = "html", align = "lrrr")
}
```

`>=5.0` testing-pattern matches the official classification. `4.3–4.9` and `<4.3` mostly do **not** — ACCESS non-takers, still classified EL.


## No post-exit collapse

```{r}
#| output: asis
emit_html(fig_html("07_exiter_percentile",
  "ILEARN ELA percentile at t (still tested) vs t+1 (ILEARN only). Primary window.",
  max_height = "420px"))
```

```{r}
#| results: asis
if (nrow(ex_ach[role == "exiter"])) {
  knitr::kable(ex_ach[role == "exiter", .(
    Band = WIDA_EXIT_BAND,
    N = prettyNum(n, big.mark = ","),
    `Pctile t` = sprintf("%.0f", ilearn_pctile_t),
    `Pctile t+1` = sprintf("%.0f", ilearn_pctile_t1),
    `Δ` = sprintf("%+.0f", ilearn_pctile_delta),
    `P(prof) t` = fmt_prob(p_proficient_t),
    `P(prof) t+1` = fmt_prob(p_proficient_t1)
  )], format = "html", align = "lrrrrrr")
}
```


## Exit-year growth sits at or above 50

ILEARN SGP at *t+1* is growth during the first year without ACCESS.

```{r}
#| output: asis
emit_html(fig_html("07_exiter_sgp",
  "Median ILEARN SGP in the exit year. Faceted by last WIDA band. Stayers = still served.",
  max_height = "400px"))
```

```{r}
#| results: asis
if (nrow(ex_sgp[role == "exiter"])) {
  knitr::kable(ex_sgp[role == "exiter", .(
    Band = WIDA_EXIT_BAND,
    N = prettyNum(n, big.mark = ","),
    `Median SGP` = sprintf("%.0f", sgp_median),
    `P(SGP < 35)` = fmt_prob(p_sgp_below_35),
    `P(SGP > 65)` = fmt_prob(p_sgp_above_65)
  )], format = "html", align = "lrrrr")
}
```


# A 50-50 proficiency cut {.center}

::: {.visually-hidden}
Section: lagged copula contrast
:::


## The recipe that gets proposed

Find the WIDA ACCESS Overall score such that a student has better than a **50-50 chance** of being ILEARN ELA-proficient **the following year**.

Read the joint through a copula $C(U_t^{\mathrm{WIDA}}, V_{t+1}^{\mathrm{ILEARN}})$. Let $v^*$ be the ILEARN proficiency cut on the paired rank scale. Invert

$$
P(V \ge v^* \mid U = u) = \tfrac{1}{2}
$$

and map $u^*$ back to a WIDA proficiency level.

Also report the empirical local-window version: the first cut $c$ at which $P(\text{proficient}_{t+1} \mid \mathrm{WIDA}_t \approx c) \ge 0.5$.

::: {.key-insight}
**`r fmt_count(n_lag)`** lagged pairs (WIDA at *t*, ILEARN at *t+1*) in the primary window.
:::


## Why we do not recommend it

::: {.warning-box}
<p class="warning-box-title">Wrong construct</p>
ILEARN proficiency is not a proxy for being ready to leave EL services. Never-EL students — native and fluent English speakers who never took ACCESS — are **not** all ILEARN-proficient. In the primary window their P(proficient) is about **`r if (nrow(lag_p)) fmt_prob(lag_p$p_never_el_proficient[1]) else "0.43"`**.
:::

If the bar is “50-50 on next year’s ELA test,” a substantial share of never-EL students fail it. Using that bar to decide English-learner status asks ACCESS to predict a content outcome that English itself does not guarantee.

We still run the recipe. It is part of the conversation. It is filed as a **contrast**.


## What the copula actually returns

```{r}
#| output: asis
emit_html(fig_html("08_lagged_p_prof_primary",
  "P(ILEARN proficient at t+1 | WIDA Overall at t). Solid = copula; points = local empirical window. Horizontal = 0.50; dotted = never-EL P(proficient).",
  max_height = "520px"))
```


## Where the 50-50 cut sits

```{r}
#| results: asis
if (nrow(lag_p)) {
  knitr::kable(data.frame(
    Quantity = c(
      "Pooled copula cut (WIDA PL at t)",
      "Pooled local-empirical cut",
      "P(prof t+1 | WIDA ≥ 4.3)",
      "P(prof t+1 | WIDA ≥ 5.0)",
      "Local P(prof t+1 | WIDA ≈ 4.3)",
      "Local P(prof t+1 | WIDA ≈ 5.0)",
      "Never-EL P(ILEARN proficient)"
    ),
    Value = c(
      fmt_num(lag_p$copula_cut[1], 2),
      fmt_num(lag_p$empirical_local_cut[1], 2),
      fmt_prob(lag_p$p_ge_4.3[1]),
      fmt_prob(lag_p$p_ge_5.0[1]),
      fmt_prob(lag_p$p_local_4.3[1]),
      fmt_prob(lag_p$p_local_5.0[1]),
      fmt_prob(lag_p$p_never_el_proficient[1])
    )
  ), format = "html", align = "lr", row.names = FALSE)
}
```

```{r}
#| output: asis
emit_html(fig_html("08_lagged_fifty_fifty_by_grade",
  "Fifty-fifty WIDA cut by ILEARN grade at t+1. Copula vs local empirical. Dashed = 4.3 and 5.0.",
  max_height = "360px"))
```

::: {.notes}
If the cut lands near 5.0, the recipe is rediscovering auto-exit by asking ACCESS to predict a hard content outcome. If it lands near 4.3, it agrees with Youden / the location table — still not a reason to treat ILEARN proficiency as the exit construct.
:::


# 4.3 versus 5.0 {.center}

::: {.visually-hidden}
Section: what this says about the exit threshold
:::


## How the pieces fit

1. ILEARN ELA is not informative about English access below WIDA **Level 3**. L1–L2 is noise in the sense of the hypothesis.
2. Signal is clearly present by the **4.3–4.9** provisional band. That is also where Youden against ILEARN proficiency peaks.
3. **5.0 auto-exit** selects students already at the top of the ILEARN distribution. It is a conservative access standard, not the first point at which ILEARN stops looking like noise.
4. Students who leave ACCESS from **5.0** hold their ILEARN location and post slightly-above-typical growth. The classification flag agrees they exited.
5. Students who disappear from ACCESS at **4.3–4.9** also hold location — but most are still classified EL. The committee-band “exiter” cell is thin and contaminated.
6. The 50-50 next-year-proficiency recipe is in the JSON and the figures. It is **not** the recommended exit rule.

::: {.key-insight}
On these files the threshold debate is about **who is allowed to leave**, not about academic collapse after ACCESS stops.
:::


## Reproducible artifacts

The pipeline writes aggregates only. Student-level pairs stay in gitignored `cache/`.

| Layer | Where |
|---|---|
| Tables | `outputs/*.csv` |
| JSON manifests | `outputs/manifests/index.json`, `results.json`, `tables/*.json` |
| Figures | `figures/*.png` (+ PDF) |
| This deck | `docs/index.html` (GitHub Pages) |

`Rscript run_all.R` rebuilds steps 00–08 and refreshes the manifests. `./docs/render.sh` re-renders this deck from `results.json`.

::: {.notes}
Pages source: Deploy from a branch → /docs. embed-resources is on, so the HTML is self-contained.
:::
